Ecosystem carbon balance in the Hawaiian Islands under different scenarios of future climate and land use change

### 2020-08-22 Manuscript/HI_Model_ms.Rmd
edited fire code and updated manuscript files

```{r fig.align = 'center', out.width = "80%", fig.cap = "Land cover classification of the seven main Hawaiian Islands, adapted from Jacobi et al (2017). Agriculture in this map combines herbaceous and woody crops, but these two crop types are treated as separate land cover classes in the simulation model. Water and Wetland land cover classes are not shown."}
knitr::include_graphics("./fig_images/fig1_HI_LC.png")
```

@@

The wildfire transition sub-model was modified from @daniel_state-and-transition_2016 by incorporating a new 21-year historical wildfire spatial database of the Hawaiian Islands (Figure S2). We used this new Hawai`i fire database to calculate historical wildfire size distribution and ignition probabilities for each unique combination of moisture zone (Figure S1), island, and state type (Figure 1) for the years 1999-2019. The number and size of fires was randomly drawn from one of these historical year-sets for each timestep and Monte Carlo realization, using burn severity probabilities from @selmants_baseline_2017. Wildfire in the low land-use scenario was sampled from the subset of historical fire years at or below the median area burned statewide from 1999-2019. The high land-use scenario sampled from historical fire years above the median area burned over the same 21-year period (Fig. S2a).

@@

 Initial carbon stocks and baseline carbon flows were estimated based on the moisture zone (Figure S1), state type, and age of each simulation cell using a lookup table derived from the Integrated Biosphere Simulator [IBIS; @foley_integrated_1996; @liu_critical_2020], a process-based dynamic global vegetation model. We initiated IBIS with minimal vegetation and simulated forward for 110 years using 30-year climate normals for the Hawaiian Islands [@giambelluca_online_2013; @giambelluca_evapotranspiration_2014]. We calibrated IBIS carbon stocks with statewide gridded datasets of soil organic carbon [@ssurgo_2016] and forest aboveground live biomass [@asner_rapid_2016]. We also calibrated gross photosynthesis in IBIS using a Hawai`i-specific gridded dataset derived from MODIS satellite imagery [@kimball_evaluating_2017]. A spatially explicit stationary growth multiplier was applied to each simulation cell to reflect local variations in net primary productivity (NPP) driven by microclimate. This spatial growth multiplier was the NPP anomaly for each cell relative the mean value for each combination of moisture zone and state type [@sleeter_effects_2019] calculated using empirical relationships between total NPP and mean annual rainfall or temperature [@schuur_productivity_2003; @del_grosso_global_2008]. Climate change impacts on carbon flows were represented by temporal growth and decay multipliers applied to each simulation cell based on statistically downscaled CMIP5 climate projections for the Hawaiian Islands under RCPs 4.5 and 8.5 [@timm_statistical_2015; @timm_future_2017]. For each RCP, temporal growth multipliers were calculated using empirical equations relating NPP to climate [@schuur_productivity_2003; @del_grosso_global_2008] and temporal decay multipliers were calculated using Q10 functions [@sleeter_effects_2019]. Transition-triggered carbon flows resulting from disturbances associated with land use change, timber harvesting, and wildfire were based on values from @don_impact_2011, @selmants_baseline_2017, and @daniel_integrating_2018.

### *Initial conditions*

@@

Tabular model output data are available from the USGS ScienceBase data repository at: https://doi.org/10.5066/P9AWLFKZ. Model input data and R code used to format input data, summarize output data, and compile this manuscript are available from the HI_Model GitHub repository at: https://github.com/selmants/HI_Model.

>>>>

```{r fig.align = 'center', out.width = "95%", fig.cap = "Land cover classification of the seven main Hawaiian Islands, adapted from Jacobi et al (2017). Agriculture in this map combines herbaceous and woody crops, but these two crop types are treated as separate land cover classes in the simulation model. Water and Wetland land cover classes are not shown."}
knitr::include_graphics("./fig_images/fig1_HI_LC.png")
```

@@

The wildfire transition sub-model was modified from @daniel_state-and-transition_2016 by incorporating a new 21-year historical wildfire spatial database of the Hawaiian Islands (Figure S2). We used this new spatial database to calculate historical wildfire size distribution and ignition probabilities for each unique combination of moisture zone (Figure S1), island, and state type (Figure 1) for the years 1999-2019. Starting in 2020, the number and size of fires was randomly drawn from one of these historical year-sets for each timestep and Monte Carlo realization, using burn severity probabilities from @selmants_baseline_2017. Wildfire in the low land-use scenario was sampled from the subset of historical fire years at or below the median area burned statewide from 1999-2019. The high land-use scenario sampled from historical fire years above the median area burned over the same 21-year period (Fig. S2a).

@@

 Initial carbon stocks and baseline carbon flows were estimated based on the moisture zone (Figure S1), state type, and age of each simulation cell using a lookup table derived from the Integrated Biosphere Simulator [IBIS; @foley_integrated_1996; @liu_critical_2020], a process-based dynamic global vegetation model. We initiated IBIS with minimal vegetation and simulated forward for 110 years using 30-year climate normals for the Hawaiian Islands [@giambelluca_online_2013; @giambelluca_evapotranspiration_2014]. We calibrated IBIS carbon stocks with statewide gridded datasets of soil organic carbon [@ssurgo_2016] and forest aboveground live biomass [@asner_rapid_2016]. We also calibrated gross photosynthesis in IBIS using a Hawai`i-specific gridded dataset derived from MODIS satellite imagery [@kimball_evaluating_2017].

 Carbon flow rates for each state type and moisture zone were estimated as the ratio of the IBIS-derived flux to the size of the originating carbon stock at each age [@sleeter_effects_2018]. A spatially explicit stationary growth multiplier was applied to each simulation cell to reflect local variations in net primary productivity (NPP) driven by microclimate. This spatial growth multiplier was the NPP anomaly for each cell relative to mean values for each combination of state type and moisture zone [@sleeter_effects_2019] calculated using empirical relationships between total annual NPP and mean annual rainfall or temperature [@schuur_productivity_2003; @del_grosso_global_2008]. Climate change impacts on carbon flows were represented by temporal growth and decay multipliers applied to each simulation cell based on statistically downscaled CMIP5 climate projections for the Hawaiian Islands under each of the two radiative forcing scenarios [RCP 4.5 and RCP 8.5; @timm_statistical_2015; @timm_future_2017]. The impact of future changes in rainfall and temperature on NPP were represented by annual growth multipliers calculated using empirical NPP models [@schuur_productivity_2003; @del_grosso_global_2008] and climate model projections of temperature and rainfall for each radiative forcing scenario. The effect of future warming on turnover rates of dead organic matter were represented by temporal decay multipliers calculated using Q10 functions and climate model temperature projections for each radiative forcing scenario. We applied a Q10 of 2.0 for wood and soil organic matter decay flows [@kurz_cbm-cfs3:_2009; @sleeter_effects_2019] and a Q10 of 2.17 for litter decay flows [@bothwell_leaf_2014]. Transition-triggered carbon flows resulting from disturbances associated with land use change, timber harvesting, and wildfire were based on values from @don_impact_2011, @selmants_baseline_2017, and @daniel_integrating_2018.

### *CO~2~ fertilization effect*

Increasing atmospheric CO~2~ concentrations stimulate leaf-level photosynthesis, potentially increasing NPP as well. However, Following @sleeter_effects_2019, we developed a set of separate scenarios designed to test the influence of a CFE on ecosystem carbon balance. We tested five $\beta$ levels ranging from 5% to 15%, which is within the range of CFEs observed in free air CO~2~ enrichment (FACE) experiments.

@@

Tabular model output data are available from a USGS ScienceBase data repository: https://doi.org/10.5066/P9AWLFKZ. Model input data and R code used to format input data, summarize output data, and compile this manuscript are available from a GitHub repository: https://github.com/selmants/HI_Model.
